---
title: "Antimicrobial concentration"
format: html
editor: visual
---

## Bacterial growth

Density-dependence can be on growth and/or death:

$$
\frac{dN}{dt} = f(N) b N - g(N) \mu N
$$

A specific example:

$$
\frac{dN}{dt} = \frac{1}{(1 + N)^\alpha}bN - (1 + N)^\beta\mu N
$$


## Antimicrobial effect

Similarly, antimicrobial effect can be on growth and/or death:

$$
\frac{dN}{dt} = f(N) b s(c) N - (g(N) \mu + k(c)) N
$$

Specific examples of $s(c)$ and $k(c)$ based on Michaelis-Menten. First the killing
rate following a Michaelis-Menten

$$
k(c) = \frac{c}{C_k + c}K
$$

where K is the maximum killing rate and $C_k$ is the antimicrobial concentration at
which half of this maximum is reached.

```{r}
k <- function(c, K, Ck) {
  K * c / (Ck + c)
}
```

For the reduction of birth:

$$
s(c) = 1 - \frac{c}{C_s + S\cdot c}S
$$

```{r}
s <- function(c, S, Cs) {
  1 - c * S / (Cs + S * c)
}
```


## Minimum inhibitory concentration

$$
bs(\mbox{mic}) = \mu + k(\mbox{mic})
$$

$$
\mbox{mic} = \frac{b + \sqrt{[(b - \mu K)^2H_s + 4\mu KbH_k]H_s}}{2\mu K} - \frac{1}{2}
$$