---
title: "Antimicrobial concentration"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Bacterial growth

Density-dependence can be on growth and/or death:

$$
\frac{dN}{dt} = f(N) b N - g(N) \mu N
$$

A specific example:

$$
\frac{dN}{dt} = \frac{1}{(1 + N)^\alpha}bN - (1 + N)^\beta\mu N
$$
```{r}
seq2 <- function(...) seq(..., le = 512)

plotl <- function(..., lwd = 8, col = 4) plot(..., type = "l", lwd = lwd, col = col)

lines2 <- function(..., lwd = 8, col = 2) lines(..., lwd = lwd, col = col)
```


```{r}
f <- function(x, alpha) {
  1 / (1 + x)^alpha
}

xs <- seq2(0, 10)
plotl(xs, f(xs, 1))
```

```{r}
g <- function(x, alpha) {
  (1 + x)^alpha
}

xs <- seq2(0, 1000)
plotl(xs, g(xs, .1))
```


```{r}
hill <- function(x, X, Y, h) {
  x2h <- x^h
  Y * x2h / (X^h + x2h)
}
```

```{r}
xs <- seq2(0, 10)
plotl(xs, hill(xs, 3, 2, 4))
lines2(xs, xs^.3)
abline(0, .2, col = 3, lwd = 8)
```


## Antimicrobial effect

Similarly, antimicrobial effect can be on growth and/or death:

$$
\frac{dN}{dt} = f(N) b s(c) N - (g(N) \mu + k(c)) N
$$

Specific examples of $s(c)$ and $k(c)$ based on Michaelis-Menten. First the killing rate following a Michaelis-Menten

$$
k(D) = \frac{KD}{C_k + D} \simeq \frac{KD}{2C_k}
$$

where K is the maximum killing rate and $C_k$ is the antimicrobial concentration at which half of this maximum is reached.

```{r}
k <- function(c, K, Ck) {
  K * c / (Ck + c)
}
```

For the reduction of birth:

$$
s(D) = \frac{C_s}{C_s + D} \simeq 1 - \frac{D}{2C_s}
$$

```{r}
s <- function(c, S, Cs) {
  1 - c * S / (Cs + S * c)
}
```

## Minimum inhibitory concentration

$$
bs(\mbox{MIC}) = \mu + k(\mbox{MIC})
$$

$$
\mbox{MIC} \simeq \frac{\lambda - \mu}{\frac{K}{2C_k} + \frac{\lambda}{2C_s}}
$$

$$
\mbox{MIC}_s = \frac{\lambda - \mu}{\mu}C_s \simeq 2\frac{\lambda - \mu}{\lambda}C_s \propto (\lambda - \mu) C_s
$$

$$
\mbox{MIC}_k = \frac{\lambda - \mu}{K - (\lambda - \mu)}C_k \simeq 2(\lambda - \mu)\frac{C_k}{K} \propto \frac{C_k}{K}\log(\mbox{R}_0)
$$

$$
\mbox{MIC} \propto \frac{\log(\mbox{R}_0)}{\mathcal{E}}
$$